---
title: "Using ChIPDBData"
author: "Yosra Berrouayel"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Using ChIPDBData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ChIPDBData)
```

# Overview

The `ChIPDBData` package provides curated ChIP-seq transcription factor target databases for use with the [`TFEA.ChIP`](https://bioconductor.org/packages/TFEA.ChIP/) package.

Each dataset contains a list of ChIP-seq experiments (e.g., from ENCODE) and their associated gene targets. These are structured as **ChIPDB** list objects and retrieved via the `getChIPDB()` accessor.

# Available Datasets

The following datasets are currently included:

- **ENCODE rE2G complete**
- ENCODE rE2G subsets filtered by score thresholds (0.25, 0.5, 0.75)
- ENCODE rE2G subsets filtered by depth thresholds (50, 100, 200, 300)
- **CREdb**
- **GeneHancer**

Use `getChIPDB()` to retrieve any of them.

# Example: Load a Dataset

```{r load-chipdb}
# Load the ENCODE dataset filtered by depth >= 200
ChIPDB <- getChIPDB("ENCODE_rE2G_300depth")
```

Each ChIPDB object is a named list with two components:

1. A character vector of Entrez Gene IDs — this is the full list of genes that could be targeted across all experiments.

2. A named list of ChIP-seq experiments, where each element corresponds to a specific experiment (typically a transcription factor in a given context), and contains integer indices pointing to the genes in the first component. These indices define the target genes for each TF ChIP-seq.

```{r explore-db}
# Inspect the top-level structure
names(ChIPDB)

# Show the first 5 Entrez IDs (the universe of genes)
ChIPDB[[1]][1:5]

# List the names of the first few ChIP-seq experiments
names(ChIPDB[[2]])[1:3]

# Show the gene indices for one experiment
ChIPDB[[2]][[1]][1:5]

# Retrieve the actual gene IDs for that experiment
ChIPDB[[1]][ ChIPDB[[2]][[1]][1:5] ]
```

# Integration with TFEA.ChIP

Once the database is loaded into the R environment as an object named `ChIPDB`, it will be automatically recognized by the TFEA.ChIP package as the active regulatory map. Important: the object must be named `ChIPDB`; otherwise, TFEA.ChIP will not detect it.

```{r eval=FALSE}
library(TFEA.ChIP)

# Example gene list
genes_of_interest <- c("TP53", "CDKN1A", "MDM2", "BAX")

# Get Entrez IDs
genes_of_interest <- GeneID2entrez( gene.IDs = genes_of_interest )

# Run TF enrichment
CM_list <- contingency_matrix( Genes.Upreg, Genes.Control ) # Generates list of contingency tables, one per dataset
results <- getCMstats( CM_list ) # Generates list of p-values and OR from association test

# Top hits
head(results)
```

# Session Info

```{r session-info}
sessionInfo()
```
